---
title: "Assignment 3 _21967657"
output: html_notebook
---

Load libraries
```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(tidyverse)
library(ggplot2)
library(readxl)
library(tidyr)
library(tools)
```

 
Task 1
```{r}
library(readr)
# Import demographic data
demographic_data <- read.csv("survey_employee_demographics.csv", na=c("","NA"))
demographic_data

# Import employment data
employment_data <- read_excel("survey_employee_employment_details.xlsx", na=c("","NA"))
employment_data

# Check the structure of the imported data frames
str(demographic_data)
str(employment_data)

# Merge the data frames based on a common column
merged_data <- merge(demographic_data,employment_data, by = "RecordID", all = TRUE)
merged_data

```

Task 2
```{r}
summary(merged_data)
```
Write at least two observations based on the summary output of the data

1.The age column ranges from a minimum value of 20 to a maximum value of 69, indicating the age distribution of the survey respondents.There are 27 missing values in the age column, suggesting potential issues with data completeness in this variable.

2.Annual Brutto Salary: The YearlyBruttoSalaryInEUR column shows a wide range of values, with the minimum salary being €10,000 and the maximum reaching an unexpectedly high value of €100 billion. However, there seems to be a data entry issue or outliers, as the mean salary is €80.28 million, which is not realistic. Additionally, there are 368 missing values in this column, which could impact salary-related analyses. Further investigation and potential data cleaning are necessary to address these anomalies.

If necessary, clean the data, rename the columns, and correct data types of the 
columns to assist with analysis.
```{r}


# Rename columns

colnames(merged_data) <- c("RecordID", "Age", "Gender", "City", "Position", "TotalYearsOfExperience", "YearsOfExperienceInGermany", "SeniorityLevel", "MainProgrammingLanguage", "OtherProgrammingLanguages", "AnnualSalaryEUR", "AnnualBonusAndStocksEUR", "AnnualSalaryOneYearAgo", "AnnualBonusStocksOneYearAgo", "NumberOfVacationDays", "EmploymentStatus", "ContractDuration", "MainLanguageAtWork", "CompanySize", "CompanyType", "JobLossDueToCovid", "ShorterWorkingWeekHours", "AdditionalSupportDueToWFH")
colnames(merged_data)

# Correct data types

merged_data$Age <- as.integer(merged_data$Age)
merged_data$TotalYearsOfExperience <- as.integer(merged_data$TotalYearsOfExperience)
merged_data$YearsOfExperienceInGermany <- as.integer(merged_data$YearsOfExperienceInGermany)
merged_data$AnnualSalaryEUR <- as.numeric(merged_data$AnnualSalaryEUR)
merged_data$AnnualBonusAndStocksEUR <- as.numeric(merged_data$AnnualBonusAndStocksEUR)
merged_data$AnnualSalaryOneYearAgo <- as.numeric(merged_data$AnnualSalaryOneYearAgo)
merged_data$AnnualBonusStocksOneYearAgo <- as.numeric(merged_data$AnnualBonusStocksOneYearAgo)
merged_data$NumberOfVacationDays <- as.integer(merged_data$NumberOfVacationDays)
merged_data$ShorterWorkingWeekHours <- as.integer(merged_data$ShorterWorkingWeekHours)
merged_data$AdditionalSupportDueToWFH <- as.numeric(merged_data$AdditionalSupportDueToWFH)
merged_data


# Identify and handle outliers (e.g., remove unrealistic salary values)

salary_threshold <- 400000  

merged_data <- merged_data %>%
  filter(AnnualSalaryOneYearAgo < salary_threshold | is.na(AnnualSalaryOneYearAgo),
        AnnualBonusStocksOneYearAgo < salary_threshold | is.na(AnnualBonusStocksOneYearAgo),
         AnnualSalaryEUR < salary_threshold | is.na(AnnualSalaryEUR),
         AnnualBonusAndStocksEUR < salary_threshold | is.na(AnnualBonusAndStocksEUR))

summary(merged_data)
merged_data





```
Task 3

```{r}
library(ggplot2)

# Visualization 1: Histogram
ggplot(merged_data, aes(x = AnnualSalaryEUR)) +
  geom_histogram(binwidth = 5000, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Yearly Brutto Salary in EUR", x = "Yearly Brutto Salary (EUR)", y = "Frequency") +
 theme_minimal()

# Visualization 2: Box Plot
ggplot(merged_data, aes(y=AnnualSalaryEUR)) +
  geom_boxplot(fill="lightgreen", color="black") +
  labs(title="Box Plot of Salary among IT Professionals",
       y="Salary (EUR)")


```
Task 4

```{r}
library(ggplot2)

# Remove rows with missing values in Gender
cleaned_data <- merged_data %>%
  filter(!is.na(Gender))
# Verify that there are no missing values in the Gender column
sum(is.na(cleaned_data$Gender))

# Box Plot
ggplot(cleaned_data, aes(x=Gender, y=AnnualSalaryEUR)) +
  geom_boxplot(fill="green", color="black") +
  labs(title="Relationship between Gender and Salary",
       x="Gender", y="Salary (EUR)")

# Bar plot to compare average salary by gender
ggplot(cleaned_data, aes(x = Gender, y = AnnualSalaryEUR)) +
           stat_summary(fun = mean, geom = "bar", fill = "green", color = "black") +
           labs(title = "Average Salary by Gender",
                x = "Gender",
                y = "Average Yearly Brutto Salary (EUR)")+
  theme_minimal()

```
Task 5

```{r}

1. # Identify and handle outliers (e.g., remove unrealistic total year experience)
summary(merged_data$total_years_experience)

total_experiance_threshold <- 100  # Setting a threshold of 100 years for total experiance

filter_total_ex <- merged_data %>%
  filter(TotalYearsOfExperience < total_experiance_threshold | is.na(TotalYearsOfExperience))

# Remove rows with missing values in total year experience
filter_total_ex<- filter_total_ex %>%
  filter(!is.na(TotalYearsOfExperience))

summary(filter_total_ex$TotalYearsOfExperience)

# Visualizing the impact of years of experience on salary
ggplot(filter_total_ex, aes(x=TotalYearsOfExperience, y=AnnualSalaryEUR)) +
  geom_point(alpha=0.7) +
  geom_smooth(method="lm", se=FALSE, color="green") +
  labs(title="Impact of Years of Experience on Salary",
       x="Years of Experience", y="Salary (EUR)")

2. # Visualizing the impact of programming languages proficiency on salary

# Remove rows with missing values in MainProgrammingLanguage
cleaned_data <- merged_data %>%
  filter(!is.na(MainProgrammingLanguage))
# Verify that there are no missing values in the MainProgrammingLanguage column
sum(is.na(cleaned_data$MainProgrammingLanguage))

# Summarize the data to get the average salary for each Main programming language
cleaned_data<- cleaned_data %>%
  group_by(MainProgrammingLanguage) %>%
  summarise(AverageSalary = mean(AnnualSalaryEUR, na.rm = TRUE))%>% arrange(desc(AverageSalary)) %>% slice(1:10)


# Print the top 10 levels
cleaned_data

bar_Languages <- ggplot(data = cleaned_data, aes(x = reorder(MainProgrammingLanguage, AverageSalary), y = AverageSalary, fill = MainProgrammingLanguage)) +
  geom_bar(stat = "identity") +
  labs(x = "MainProgrammingLanguage", y = "Salary(EUR)", fill = "MainProgrammingLanguage") +
  ggtitle("Impact of MainProgrammingLanguage on Average Income") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + coord_flip()
  theme_minimal()

print(bar_Languages)


3.  # Visualizing the impact of contract duration on salary

# Remove rows with missing values in Contract Duration
filter_contract_duration <- merged_data %>%
  filter(!is.na(ContractDuration))

filter_contract_duration <- filter_contract_duration %>% filter(ContractDuration != 0)

# box plot for contract duration vs Salary
box_contract <- ggplot(data = filter_contract_duration, aes(x = ContractDuration, y = AnnualSalaryEUR, fill = ContractDuration)) +
  geom_boxplot() +
  labs(x = "Contract Duration", y = "Yearly Brutto Salary (EUR)") +
  ggtitle("Annual Salary based on Contract Duration") +
  theme_minimal() +
  theme(legend.position = "none")

print(box_contract)

4.  # Visualizing the impact of seniority level on salary


# Remove rows with missing values in Seniority Level and AnnualSalaryEUR
cleaned_data <- merged_data %>%
  filter(!is.na(SeniorityLevel), !is.na(AnnualSalaryEUR))

# Verify that there are no missing values in the Seniority Level column
sum(is.na(cleaned_data$SeniorityLevel))

# Capitalize first letter of each category using str_to_title to remove duplicate same category
cleaned_data$SeniorityLevel <- str_to_title(cleaned_data$SeniorityLevel)

# Reorder SeniorityLevel based on average AnnualSalaryEUR
cleaned_data <- cleaned_data %>%
  mutate(SeniorityLevel = fct_reorder(SeniorityLevel, AnnualSalaryEUR, .fun = mean, .desc = TRUE))

# Bar plot to compare average salary by seniority level
ggplot(cleaned_data, aes(y = SeniorityLevel, x = AnnualSalaryEUR)) +
  stat_summary(fun = mean, geom = "bar", fill = "green", color = "black") +
  labs(title = "Average Salary based on Seniority Level",
       x = "Average Yearly Salary (EUR)",
       y = "Seniority Level") +
  theme_minimal()


```








